Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-03-26 paper/main.tex
first go at marc's second suggestions

\title{A meta analysis of tournaments and an evaluation of performance in the
Iterated Prisoner's Dilemma.}
\author{Nikoleta E. Glynatsi, Vincent A. Knight}
\date{}

@@

\begin{abstract}
The Iterated Prisoner's Dilemma has been used for decades as a model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for decades. The results of the literature,
however, have been relying on a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments,
casting doubt on the generalizability of results from statistical perspective.
This manuscript evaluates \numberofstrategies strategies in more than 40000
tournaments, presents the top performing strategies, and analyzes their
salient features.
The outcomes support that a strategy with a theory of mind should adjust its
cooperations based on the mean/median cooperator of its environment to be
successful; the adjustments are distinct for different tournament types.
\end{abstract}

@@

In the 80's, Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b}. The contestants were strategies submitted in the form of computer
code. They competed against all other entries, a copy of themselves and a random
strategy. The winner was decided on the average score a strategy achieved. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of its opponent.
Due to the strategy's strong performance in both tournaments, and moreover in a
series of evolutionary experiments~\cite{Axelrod1981}, Tit For Tat was often
claimed to be the most robust basic strategy in the IPD.

However, further research proved that the strategy had weakness, and more
specifically, it was shown that the strategy suffered in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984}. This was
mainly due to the strategy's lack of generosity and contrition. The strategy was
quick to punish a defection, and in a noisy environment it could lead to a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced and became the new
protagonists of the game. These include Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.

In 2004, a $20^{\text{th}}$ Anniversary Iterated Prisoner Dilemma Tournament
took place with 233 entries. This time the winning strategy was not designed on
a reciprocity based approach but on a mechanism of
teams~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}. A team
from Southampton University took advantage of the fact that a participant was
allowed to submit multiple strategies. They submitted a total of 60 strategies
that could recognise each other and collude to increase one members score.
This resulted in three of the strategies ranking in the top spots. The
performance of the Southampton University team received mixed attention, though
they had won the tournament as stated in~\cite{us_blog} "technically this
strategy violates the spirit of the Prisoner's Dilemma, which assumes that the
two prisoners cannot communicate with one another".

Another set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their superiority in tournament settings
have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments come in a
number of variations including tournaments with noise, probabilistic match
length and both noise and probabilistic match length.
The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

@@

For the purposes of this manuscript a data set containing results of IPD
tournaments has been generated and is available at~\cite{data}. This was done using the
open source package Axelrod-Python library~\cite{axelrodproject} (APL), and more specifically,
version 3.0.0. APL allows for different types of IPD computer
tournaments to be simulated whilst containing a list of over 180 strategies.
Most of these are strategies described in the literature with a few exceptions
being strategies that have been contributed specifically to the package. This
paper makes use of \numberofstrategies strategies implemented in version 3.0.0. A
list of the strategies is given in the Appendix~\ref{app:list_of_players}.
Although APL features several tournament types, this work considers
standard, noisy, probabilistic ending and noisy probabilistic ending
tournaments.

@@

The source code for the data collection, as well as the source code for
the analysis, which will be discussed in the following sections, have been written
following best practices~\cite{Aberdour2007, Benureau2018}. It has been packaged
and is available here. %TODO archive code

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.60\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.34.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

The top 15 strategies for each tournament type based on \(\bar{r}\) are given in
Table~\ref{table:top_performances}. The data collection process was design such
as that the probabilities of noise and ending of the match to vary between 0 and
1. However, the most common values for these probabilities in the literature are
values less than 0.1. Thus,
Table~\ref{table:top_performances} also includes the top 15 strategies in noisy
tournaments with \(p_n < 0.1\) and probabilistic ending tournaments with \(p_e <
0.1\).

\newcolumntype{g}{>{\columncolor{Gray}}l}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{\textwidth}{!}{
        \input{top_perfomances.tex}
    }
\end{center}
\caption{Top performances for each tournament type based on $\bar{r}$.}
\label{table:top_performances}
\end{table}

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in APL in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, and Fool Me Once, are
strategies not from the literature but from the APL. DoubleCrosser
is a strategy that makes use of the number of turns because it is set to defect on
the last two rounds. The strategy was expected to not perform as well in
tournaments where the number of turns is not specified. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is a strategy specifically designed for noisy environments, however, it ranks
highly in standard tournaments as well.

Figure~\ref{fig:std_results} gives the distributions of $r$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median, of the top 15 strategies, is at 0.075. This indicates that the top ranked
strategies perform well in any given standard tournament, despite the opponents and
the number of turns.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_standard.pdf}
    \caption{$r$ distributions of top 15 strategies in standard tournaments.}\label{fig:std_results}
\end{figure}

In the case of noisy tournaments with \(p_n < 0.1\) the top performed strategies
include strategies specifically designed for noisy tournaments such as DBS,
Evolved FSM 16 Noise 05, Evolved ANN 5 Noise 05, PSO Gambler 2 2 2 Noise 05 and
Omega Tit For Tat~\cite{kendall2007iterated}, strategies which performed well in
standard tournaments such as BackStabber, DoubleCrosser, Evolved ANN, FSM and
HMM and deterministic strategies such as Spiteful Tit For Tat~\cite{prison},
Level Punisher~\cite{Eckhart2015}, Eugine Nier~\cite{lesswrong}. The \(r\) distributions
of these strategies are given in Figure~\ref{fig:noise_results} where it is evident
that these strategies overall performed well in noisy tournaments.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/noise_performance_subset.pdf}
    \caption{$r$ distributions of top 15 strategies in 1151 noisy tournaments with \(p_n < 0.1\).}\label{fig:noise_results}
\end{figure}

In comparison, the top ranked strategies in noisy environments when \(p_n\in [0,
1]\) include strategies which decide their actions based on the cooperations to
defections ratio, such as ShortMem~\cite{Andre2013},
Grumpy~\cite{axelrodproject} and e~\cite{axelrodproject}, and the Retaliate
strategies which are designed to defect if the opponent has tricked them more
often than x\% of the times that they have done the same. The overall
performance of these strategies is bimodal,
Figure~\ref{fig:noise_overall_results}, indicating that although the top
ranked strategies mainly performed well, there are several tournaments that they
ranked in the bottom half. The bimodality of the \(r\) distributions can be
explained by Figure~\ref{fig:effect_of_noise} which gives the \(r\)
distributions for the top 6 strategies over the noise probability \(p_n\).
It is evident that the highly ranked strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_noise.pdf}
    \caption{$r$ distributions of top 15 strategies in noisy tournaments.}\label{fig:noise_overall_results}
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.6\textwidth]{../images/probend_performance_subset.pdf}
    \caption{\(r\) distributions for best performed strategies in 1139 probabilistic ending
    tournaments with \(p_e < 0.1\).}
    \label{fig:probend_results}
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_probend_noise.pdf}
    \caption{\(r\) distributions for best performed strategies in noisy
    probabilistic ending tournaments.}
    \label{fig:noisy_probend_results}
\end{figure}

\begin{table}[!htbp]
    \centering
    \resizebox{.27\textwidth}{!}{
    \input{performance_probend_noise_subset_table.tex}}
    \caption{Top performances in 117 probabilistic ending tournaments with \(p_e < 0.1\) and \(p_n < 0.1\).}
    \label{table:subset_probend_noise_result}
\end{table}

Up till now, the performances of the \numberofstrategies strategies have been
evaluated for individual tournament types. The distributions of \(r\) for the
tournament types indicate that for standard, probabilistic ending and noisy
tournaments with \(p_n <0.1\) successful strategies do exist. For these settings
the top 15 strategies frequently ranked in the top spots with only a few
exceptions. The behaviour of the winning strategies in probabilistic ending
tournaments varied when \(p_e\) was restricted to 0.1. For larger values of
\(p_e\) the dominant strategies were defecting ones and for smaller values the
dominant strategies were evolved strategies and strategies based on teams.
Similarly the behaviour
of the dominant strategies varied when the probability of noisy was restricted.
In noisy tournaments with \(p_n < 0.1\) the dominant strategies were strategies
specifically trained for noise tournaments, and the performed well in most of the
tournaments they participated. However, in noisy tournaments with \(p_n \in [0, 1]\)
the performances of the top ranked strategies were bimodal. Indicating that with
high values of noisy a successful set of strategies does not exist.

The data set considered in this work, described in
Section~\ref{section:data_collection}, contains a total of
\numberofalltournaments tournament results. For this part of the manuscript the
strategies are ranked based on the median normalised rank they achieved over the
entire data set. The top 15 strategies are given in
Table~\ref{table:overall_results} and their normalised rank distributions are
given in Figure~\ref{fig:overall_results}.

\begin{table}[!htbp]
    \centering
    \resizebox{.27\textwidth}{!}{
    \input{performance_merged_table.tex}}
    \caption{Top performances over all the tournaments}\label{table:overall_results}
\end{table}

The top ranks include strategies that have been previously mentioned. The
set of Retaliate strategies occupy the top spots followed by BackStabber
and DoubleCrosser. The distributions of the Retaliate strategies have no
statistical difference. Thus, in an IPD tournament where the type is not
specified, playing as any of the Retaliate strategies will have same the result.
DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger. Overall the top 15 strategies are fundamentally different. Some are cheaters,
some are complex, others are simple deterministic strategies and strategies based
on teams. The results of \numberofalltournaments tournaments used in this work imply the following:
they is not a single type of strategy performs well in any IPD interaction.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_merged.pdf}
    \caption{\(r\) distributions for best performed strategies in the data set~\cite{data}.}
    \label{fig:overall_results}
\end{figure}

This section presented the winning strategies in a series of IPD tournaments. In
standard tournaments the top spots were dominated by complex strategies that had
been trained using reinforcement learning techniques. In noisy environments where
the a noise probability strictly less than 0.1 was considered the successful
strategies were strategies based on the behaviour of many strategies and strategies
specifically designed for noisy tournaments. In
probabilistic ending tournaments most of the highly ranked strategies were defecting
strategies and trained finite state automata, all by the authors of~\cite{Ashlock2006, Ashlock2014}.
These strategies ranked high due to their performance in tournaments where the probability of the game ending after
each turn was bigger than 0.1. In probabilistic tournaments with \(p_e\) less than 0.1
the highly ranked strategies were trained and Meta strategies.
Finally the performance of all \numberofstrategies
strategies over the \numberofalltournaments tournaments in this manuscript was
assessed on \(\bar{r}\). The top ranked strategies were a mixture of behaviours
that did well in standard tournaments, tournaments with noise and a
few Meta strategies.

The results of this section imply that successful strategies for specific
settings exist for an IPD tournament. The top ranked strategies in both standard
tournaments and tournaments with probabilistic ending, managed to rank in the
top 10\% of the tournament most of the times.
Strategies in noisy
environments demonstrated that no strategy can be consistently
successful, expected if the value of noise is constrained to less than a 0.1.
Trained strategies, meta strategies and some deterministic defecting strategies
have demonstrated good performance over several tournament types, however,
there exists no one single strategy which appears to be dominant in an IPD
competition. The aim of the next section is to understand which are
the factors that made these strategies successful, in each setting separately
but also overall.

@@

\textbf{Evaluation based on clustering and random forest.}

The second method to evaluate the features importance in a strategy's success
is a combination of a clustering task and a random forest algorithm.
Initially the performances are clustered into different clusters
based on them being successful or not. The performances are clustered into
successful and unsuccessful clusters based on 4 different approaches.
More specifically:

\begin{itemize}
    \item \textbf{Approach 1:} The performances are divided into two clusters based
    on whether their performance was in the top \(5\%\) of their respective tournaments.
    Thus, whether \(r\) was smaller or larger than 0.05.
    \item \textbf{Approach 2:} The performances are divided into two clusters based
    on whether their performance was in the top \(25\%\) of their respective tournaments.
    Thus, whether \(r\) was smaller or larger than 0.25.
    \item \textbf{Approach 3:} The performances are divided into two clusters based
    on whether their performance was in the top \(50\%\) of their respective tournaments.
    Thus, whether \(r\) was smaller or larger than 0.50.
    \item \textbf{Approach 4:} The performances are clustered based on their normalised rank and
    their median score by a \(k-\)means algorithm~\cite{Arthur2007}. The number of
    clusters is not deterministically chosen but it is based on the silhouette
    coefficients~\cite{Rousseeuw1987}.
\end{itemize}

Once the performances have been assigned to a cluster for each approach a
random forest algorithm~\cite{breiman2001} is applied. The problem is a supervised
problem where the random forest algorithm predicts the cluster to which a performance
has been assigned to using the features of Table~\ref{table:manual_features}.
The random
forest models are trained on a training set of 70\% of the tournaments results.
The accuracy of each model based on $R^2$ and the number of clusters for each
tournament type (because in the case of Approach 4 it is not
deterministically chosen) are given by Table~\ref{table:accuracy_random_forest}.
The out of the bag error (OOB)~\cite{hastie2005} has also been calculated. The
models fit well, and a high value of both the accuracy measures on the test data
and the OOB error indicate that the model is not over fitting.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.9\textwidth}{!}{
        \begin{tabular}{lccccc}
    \toprule
    Tournament type & Clustering Approach & Number of clusters & $R^2$ training data &  $R^2$ test data  & $R^2$ OOB score\\
    \midrule
    standard  & Approach 1  & 2 & 0.998831  & 0.987041    & 0.983708 \\
              & Approach 2  & 2 & 0.998643  & 0.978626    & 0.969202 \\
              & Approach 3  & 2 & 0.998417  & 0.985217    & 0.976538 \\
              & Approach 4  & 2 & 0.998794  & 0.990677    & 0.982959 \\
    \midrule
    noisy     & Approach 1  & 2 & 0.997786  & 0.972229    & 0.968332\\
              & Approach 2  & 2 & 0.997442  & 0.963254    & 0.955219\\
              & Approach 3  & 2 & 0.997152  & 0.953164    & 0.940528\\
              & Approach 4  & 3 & 0.996923  & 0.950728    & 0.935444\\
    \midrule
    probabilistic ending & Approach 1 & 2 & 0.997909   & 0.981490  & 0.978120 \\
                         & Approach 2 & 2 & 0.997883   & 0.973492  & 0.967150 \\
                         & Approach 3 & 2 & 0.990448   & 0.890068  & 0.875822 \\
                         & Approach 4 & 2 & 0.999636   & 0.995183  & 0.992809 \\
    \midrule
    noisy probabilistic ending & Approach 1  & 2 & 0.995347 & 0.957846   & 0.952353\\
                               & Approach 2  & 2 & 0.992813 & 0.909346   & 0.898613\\
                               & Approach 3  & 2 & 0.990579 & 0.824794   & 0.806540\\
                               & Approach 4  & 4 & 0.989465 & 0.841652   & 0.824052\\
    \midrule
    over \numberofalltournaments tournaments   & Approach 1 & 2 & 0.997271& 0.972914& 0.969198 \\
                                               & Approach 2 & 2 & 0.996323& 0.951194& 0.940563 \\
                                               & Approach 3 & 2 & 0.993707& 0.906941& 0.891532 \\
                                               & Approach 4 & 3 & 0.993556& 0.913335& 0.898453 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Accuracy metrics for random forest models.}
    \label{table:accuracy_random_forest}
\end{table}

The importance that the features of Table~\ref{table:manual_features} had on
each random forest model are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the features that are highlighted by the importance are inline with
the correlation results. Moreover, the smoothing parameter \(k\) appears to no
have a significant effect either. The most important features based on the
random forest analysis were $C_{r} / C_{median}$ and $C_r / C_{mean}$.

>>>>

\title{Properties of winning Iterated Prisoner's Dilemma strategies.}
\author{Nikoleta E. Glynatsi, Vincent A. Knight, Marc Harper}
\date{}

@@

\begin{abstract}
From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of Iterated Prisoner's Dilemma strategies for decades. The results of the literature,
however, have been relying on a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments,
casting doubt on the generalizability of results from statistical perspective.
This manuscript evaluates \numberofstrategies strategies in 45686
tournaments, presents the top performing strategies, and analyzes their
salient features. The results of the tournaments imply that there is not a single
type of strategy that performs well in any Iterated Prisoner's Dilemma interaction.
There are several properties, however, that heavily influence the best performing
strategies: be nice, be provocable, be a little envious and be clever.
\end{abstract}

@@

In the 80's, Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b}. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of its opponent.
Axelrod concluded that the strategy's robustness was due to four different
properties:

\begin{itemize}
    \item Do not be envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection
    \item Do not be too clever
\end{itemize}

As a result of the strategy's strong performance in both tournaments, and moreover in a
series of evolutionary experiments~\cite{Axelrod1981}, Tit For Tat was often
claimed to be the most robust basic strategy in the IPD.

Further research proved that the strategy had weakness, and more
specifically, it was shown that the strategy suffered in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984}. This was
mainly due to the strategy's lack of generosity and contrition. The strategy was
quick to punish a defection, and in a noisy environment it could lead to a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced and became the new
protagonists of the game. These include Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.
Several works extended the reciprocity based approach which has led to new
strategies. For example Gradual~\cite{Beaufils1997} which was constructed to
have the same qualities as those of Tit for Tat except one,
\textit{Gradual} had a memory of the game since the beginning of it. Gradual
recorded the number of defections by the opponent and punished them with a
growing number of defections. It would then enter a calming state in which it
would cooperates for two rounds. In a tournament of 12 strategies, including
both Tit for Tat and Pavlov, Gradual managed to outperformed them all. A
strategy with the same intuition as Gradual is \textit{Adaptive Tit for
Tat}~\cite{tzafestas-2000a}. Adaptive Tit for Tat does not keep a permanent
count of past defections, it maintains a continually updated estimate of the
opponent’s behaviour, and uses this estimate to condition its future actions. In
the exact same tournament as in~\cite{Beaufils1997} with now 13 strategies Adaptive
Tit for Tat ranked first.
A set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their superiority in tournament settings
have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments come in a
number of variations including tournaments with noise, probabilistic match
length and both noise and probabilistic match length.
The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The outcomes of our
work reinforce the discussion started by Axelrod:

\begin{itemize}
    \item Do not be envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection
    \item Do not be too clever
\end{itemize}

Furthermore, the data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

@@

For the purposes of this manuscript a data set containing results of IPD
tournaments has been generated and is available at~\cite{data}. This was done using the
open source package Axelrod-Python library~\cite{axelrodproject} (APL), and more specifically,
version 3.0.0. APL allows for different types of IPD computer
tournaments to be simulated whilst containing a large list of strategies.
Most of these are strategies described in the literature with a few exceptions
being strategies that have been contributed specifically to the package. This
paper makes use of \numberofstrategies strategies implemented in version 3.0.0. A
list of the strategies is given in the Appendix~\ref{app:list_of_players}.
Although APL features several tournament types, this work considers
standard, noisy, probabilistic ending and noisy probabilistic ending
tournaments.

@@

The source code for the data collection, as well as the source code for
the analysis, which will be discussed in the following sections, have been written
following best practices~\cite{Aberdour2007, Benureau2018}
and is available here. %TODO archive code

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.60\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. Lower values of \(r\)
    correspond to better performances. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.34.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

The top 15 strategies for each tournament type based on \(\bar{r}\) are given in
Table~\ref{table:top_performances}. The data collection process was design such
as that the probabilities of noise and ending of the match to vary between 0 and
1. However, commonly used values of these probabilities are
values less than 0.1. Thus,
Table~\ref{table:top_performances} also includes the top 15 strategies in noisy
tournaments with \(p_n < 0.1\) and probabilistic ending tournaments with \(p_e <
0.1\).

\newcolumntype{g}{>{\columncolor{Gray}}l}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{\textwidth}{!}{
        \input{top_perfomances.tex}
    }
\end{center}
\caption{Top performances for each tournament type based on $\bar{r}$. The results
of each type are based on 11420 unique tournaments of each type. The results
for noisy tournaments with \(p_n < 0.1\) are based on 1151 tournaments, and 
for probabilistic ending tournaments with \(p_e < 0.1\) on 1139. The top ranks
indicate that trained strategies perform well in a variety of environments,
but so do simple deterministic strategies. The normalised medians are close to 0
for most environments, except environments with noise not restricted to 0.1.
These two tournaments have the highest medians, implying that strategies from
the collection of this work can not perform well in environments with such high
values of noise.}
\label{table:top_performances}
\end{table}

Distributions of these top ranked strategies are given by Figure~\ref{fig:r_distributions}.
\newpage

\begin{figure*}[!htbp]
    \centering
    \begin{subfigure}{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../images/performance_standard.pdf}
        \caption{$r$ distributions in standard tournaments.}\label{fig:std_results}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../images/noise_performance_subset.pdf}
        \caption{$r$ distributions of top 15 strategies in 1151 noisy tournaments with \(p_n < 0.1\).}\label{fig:noise_results}
    \end{subfigure}
    \vskip\baselineskip
    \begin{subfigure}{0.475\textwidth}
        \centering 
        \includegraphics[width=\textwidth]{../images/performance_noise.pdf}
        \caption{$r$ distributions of top 15 strategies in noisy tournaments.}\label{fig:noise_overall_results}
    \end{subfigure}
    \quad
    \begin{subfigure}{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../images/probend_performance_subset.pdf}
        \caption{\(r\) distributions for best performed strategies in 1139 probabilistic ending
        tournaments with \(p_e < 0.1\).}
        \label{fig:probend_results}
    \end{subfigure}
    \caption{\(r\) distributions of the top 15 strategies in different environments. A lower value of
    \(\bar{r}\) corresponds to a more successful performance. A strategy's \(r\) distribution skewed towards zero
    indicates that the strategy ranked highly in most tournaments it participated in.}\label{fig:r_distributions}
\end{figure*}

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in APL in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, BackStabber and Fool Me Once, are
strategies not from the literature but from the APL. DoubleCrosser is an extension
of BackStabber and both strategies make use of the number of turns because they are
set to defect on the last two rounds. It should be noted that these
strategies can be characterised as "cheaters". The source code of the strategies
allows them to known the number of turns in a match (if they are specified).
These strategies were expected to not perform as well in
tournaments where the number of turns is not specified. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is a strategy specifically designed for noisy environments and it ranks
highly in standard tournaments as well.
Figure~\ref{fig:std_results} gives the distributions of $r$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median, of the top 15 strategies, is at 0.075. This indicates that the top ranked
strategies perform well in any given standard tournament, despite the opponents and
the number of turns.

In the case of noisy tournaments with \(p_n < 0.1\) the top performed strategies
include strategies specifically designed for noisy tournaments such as DBS,
Evolved FSM 16 Noise 05, Evolved ANN 5 Noise 05, PSO Gambler 2 2 2 Noise 05 and
Omega Tit For Tat~\cite{kendall2007iterated}, strategies which performed well in
standard tournaments such as BackStabber, DoubleCrosser, Evolved ANN, FSM and
HMM and deterministic strategies such as Spiteful Tit For Tat~\cite{prison},
Level Punisher~\cite{Eckhart2015}, Eugine Nier~\cite{lesswrong}. The \(r\) distributions
of these strategies are given in Figure~\ref{fig:noise_results} where it is evident
that these strategies overall performed well in noisy tournaments.
In comparison, the top ranked strategies in noisy environments when \(p_n\in [0,
1]\) include strategies which decide their actions based on the cooperations to
defections ratio, such as ShortMem~\cite{Andre2013},
Grumpy~\cite{axelrodproject} and e~\cite{axelrodproject}, and the Retaliate
strategies which are designed to defect if the opponent has tricked them more
often than x\% of the times that they have done the same. The overall
performance of these strategies is bimodal,
Figure~\ref{fig:noise_overall_results}, indicating that although the top
ranked strategies mainly performed well, there are several tournaments that they
ranked in the bottom half. The bimodality of the \(r\) distributions can be
explained by Figure~\ref{fig:effect_of_noise} which gives the \(r\)
distributions for the top 6 strategies over the noise probability \(p_n\).
It is evident that the highly ranked strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly.

@@

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.55\textwidth}{!}{
    \input{performance_probend_noise_subset_table.tex}}
    \caption{Top performances in 117 probabilistic ending tournaments with \(p_e < 0.1\) and \(p_n < 0.1\).}
    \label{table:subset_probend_noise_result}
\end{table}

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../images/performance_probend_noise.pdf}
        \caption{\(r\) distributions for best performed strategies in noisy
        probabilistic ending tournaments.}
        \label{fig:noisy_probend_results}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../images/performance_merged.pdf}
        \caption{\(r\) distributions for best performed strategies in the data set~\cite{data}.}
        \label{fig:overall_results}
    \end{subfigure}
    % \caption{\(r\) distributions of the top 15 strategies in different environments. A lower value of
    % \(\bar{r}\) corresponds to a more successful performance. A strategy's \(r\) distribution skewed towards zero
    % indicates that the strategy ranked highly in most tournaments it participated in.}\label{fig:r_distributions}
\end{figure}

This section presented the winning strategies in a series of IPD tournaments. In
standard tournaments the top spots were dominated by complex strategies that had
been trained using reinforcement learning techniques. In noisy environments where
the a noise probability strictly less than 0.1 was considered the successful
strategies were strategies based on the behaviour of many strategies and strategies
specifically designed for noisy tournaments. In
probabilistic ending tournaments most of the highly ranked strategies were defecting
strategies and trained finite state automata, all by the authors of~\cite{Ashlock2006, Ashlock2014}.
These strategies ranked high due to their performance in tournaments where the probability of the game ending after
each turn was bigger than 0.1. In probabilistic tournaments with \(p_e\) less than 0.1
the highly ranked strategies were trained and Meta strategies.
Finally the performance of all \numberofstrategies
strategies over the \numberofalltournaments tournaments in this manuscript was
assessed on \(\bar{r}\). The top ranked strategies were a mixture of behaviours
that did well in standard tournaments, tournaments with noise and a
few Meta strategies.
The results of this section imply that successful strategies for specific
settings exist for an IPD tournament. The top ranked strategies in both standard
tournaments and tournaments with probabilistic ending, managed to rank in the
top 10\% of the tournament most of the times.
Strategies in noisy
environments demonstrated that no strategy can be consistently
successful, expected if the value of noise is constrained to less than a 0.1.
Trained strategies, meta strategies and some deterministic defecting strategies
have demonstrated good performance over several tournament types, however,
there exists no one single strategy which appears to be dominant in an IPD
competition. The aim of the next section is to understand which are
the factors that made these strategies successful, in each setting separately
but also overall.

Be envious and be clever.

@@

A multivariate linear regression is fitted to model the relationship of
Table's~\ref{table:manual_features} and the normalised rank. Based on the
correlation matrices. Appendix~\ref{app:correlations} the depended variables are
correlated. The median cooperating ration \(C_median\) and \(C_r/ C_median\) are highly
correlated with \(C_mean\) and \(C_r / C_mean\). So these features are dropped from
the analysis. The \(C_r\) is also correlated and drop. Finally for some environments
the \(C_max\) and \(C_min\) are correlated and dropped for the models of those types.
More specifically, the list of features their p values and \(p-\)values of
each feature as well as their coefficients are given by
Table~\ref{table:linear_regression}. The coefficient of determination, \(R\)
squared, for each model is also reported in Table~\ref{table:linear_regression}.
